Isomorphic Interpreters from Logically Reversible Abstract Machines

### 2012-03-16 iso-int/rc-iso-int.tex
revised

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
% Previous work on {{langRevT}} \cite{rc2011,infeffects} studied its
% properties and various constructions in it. However, for a
% computational model, there has been little focus on actually writing
% programs directly in {{langRevT}}. This paper aims to remedy that by
% presenting a systematic technique of expressing logically reversible
% small-step abstract machines in {{langRevT}}. In other words, this
% paper shows that once we have a logically reversible machine in a
% notation of our choice, expressing the machine as an isomorphic
% interpreter in {{langRevT}} can be done systematically and does not
% present any significant conceptual difficulties. This gives us a means
% of developing large {{langRevT}} programs with the ease of reasoning
% at the level of a conventional small-step semantics.  We develop
% several simple interpreters over numbers and addition, move onto tree
% traversal and finish with an interpreter for {{langRevT}} expressed in
% {{langRevT}}.
  In our previous work, we developed a reversible programming language and
  established that every computation in it is a (partial) isomorphism that is
  reversible and that preserves information. The language is founded on type
  isomorphisms that have a clear categorical semantics but that are awkward
  as a notation for writing actual programs, especially recursive
  programs. This paper remedies this aspect by presenting a systematic
  technique for developing a large and expressive class of reversible
  recursive programs, that of logically reversible small-step abstract
  machines. In other words, this paper shows that once we have a logically
  reversible machine in a notation of our choice, expressing the machine as
  an isomorphic interpreter can be done systematically and does not present
  any significant conceptual difficulties. Concretely, we develop several
  simple interpreters over numbers and addition, move on to tree traversals,
  and finish with a meta-circular interpreter for our reversible
  language. Generally, this gives us a means of developing large reversible
  programs with the ease of reasoning at the level of a conventional
  small-step semantics.
\end{abstract}

@@

In this paper, we investigate the practicality of writing recursive programs
in {{langRevT}}. Specifically, we assume that we are giving some reversible
recursive algorithm expressed as a \emph{reversible abstract machine}, and we
show via a number of systematic steps, how to develop a {{langRevT}} program
from that abstract machine. Our choice of starting from reversible abstract
machines is supported by the following observations:
\begin{itemize}
\item it is an interesting enough class of reversible programs: researchers
  have invested the effort in manually designing such machines, e.g., the
  SE(M)CD machine~\cite{Kluge:1999:SEMCD}, and the SECD-H~\cite{lorenz};
\item every reversible interpreter can be realized as such a machine: this
  means that our class of programs includes self-interpreters which are
  arguably a measure of the strength of any reversible
  language~\cite{Yokoyama:2008:PRP,Axelsen:2011:RPC:1987171.1987176,Yokoyama:2007:RPL:1244381.1244404};
\item general recursive programs can be systematically transformed to
  abstract machines: the technique is independent of reversible programs and
  consists of transforming general recursion to tail recursion and then
  applying \emph{fission} to split the program into a driver and a small-step
  machine~\cite{Danvy:2008:RRN:1813347.1813350,Rendel:2010:ISD:1863523.1863525}.
\end{itemize}

@@

\paragraph*{Graphical Language.}
The syntactic notation above is often obscure and hard to read.  Following
the tradition established for monoidal
categories~\cite{springerlink:10.1007/978-3-642-12821-94}, we present a
graphical language that conveys the semantics of the language.  The general
idea of the graphical notation is that combinators are modeled by ``wiring
diagrams'' or ``circuits'' and that values are modeled as ``particles'' or
``waves'' that may appear on the wires. Evaluation therefore is modeled by
the flow of waves and particles along the wires.

@@

\item Operations {{fold}} and {{unfold}} are specific to the recursive
  data type and are drawn as triangular boxes. For instance,
  Sec. \ref{sec:bounded} shows {{fold}}/{{unfold}} isomorphisms for
  numbers, {{nat=@@x.(1+x)}}.

\item The {{trace+}} operation is drawn as a looped circuit, where the
  traced type {{b1}} is shown as flowing backward. The semantics and
  interpretation of {{trace}} are dealt with in detail in previous
  work (see Sec. 4 of \cite{James:2012:IE:2103656.2103667}).

@@

Our goal is to implement this abstract machine in {{langRevT}}: We
start by writing a combinator {{c : nat * nat <-> nat * nat}} that
executes a `small step'. Here {{nat}} is an abbreviation for the
recursive type {{@@x.1+x}}. The combinator, when iterated, should map
{{(n,0)}} to {{(0,n)}} by precisely mimicking the steps of the
abstract machine. Let us analyze this desired combinator {{c}} in
detail starting from its interface:

@@

The triangle on the left denotes the {{unfold}} isomorphism, which
works as follows: If the number {{n}} is zero, the output is the top
branch which has type {{1}}. If the number was non-zero, the output is
on the bottom branch and has value {{n-1}}. The triangle on the right
is the dual {{fold}} isomorphism.  Let us examine all the possible
machine states by introducing distribution operations and any required
swaps:

@@

The construction above already gives us some insight into how we can
implement our machine's one reduction step. We want the output {{n'}} to be
{{n+1}} where~{{n}} is the first component of the input and similarly the
output {{m'}} to be {{m+1}} where {{m}} is the second component of the input.
We can express these equalities by connecting wires, i.e., by connecting the
{{n-1}} wire to {{n'}} and the {{m}} wire to {{m'-1}}:

@@

The branch labeled {{((), m)}} corresponds to the machine state {{<0,m>}} and
hence looks like the stop state of the machine. Similarly the branch labeled
{{((), n')}} corresponds to the start state of the machine. We are close to
the completion of the interpreter at this point. The last step in the
construction involves introducing a {{trace}} operation for iteration of
reduction steps and exposing the start and stop states. We do this by
``sliding'' the two sections of the diagram as shown below:

@@

The introduced {{trace}} feeds back related machine states. The
exposed start state essentially requires the initial value of {{n}} to
start the machine and the exposed stop state returns any result values
that have been computed.

@@

A closer look however reveals that the machine defines a
\emph{partial} isomorphism: not all valid final states can be mapped
to valid start states. Indeed consider the configuration {{<0,2,3>}}
which is a valid final state. Going backwards, the transitions start
as follows {{<0,2,3> |--> <1,1,2> |--> <2,0,1>}} at which point, the
machine gets stuck at a state that is not a valid start state.

\paragraph*{Stuck States.}
All abstract machines, not just those written in {{langRevT}}, have to
deal with stuck states. There are several standard ways of dealing
with them:
\begin{itemize}
\item \emph{Use better types.}  One could augment {{langRevT}} with a
  richer type system that distinguishes non-zero numbers from those
  that can be zero, thereby eliminating the {{sub1 0}} situation
  entirely. Similarly, in the meta-circular interpreter in
  Sec. \ref{sec:langrevt-int} one could use GADTs to eliminate stuck
  states.

\item \emph{Divergence.}  Another standard approach in dealing with
  stuck states is to make the machine diverge or leave the output
  undefined/unobservable in some way. The {{add1}} combinator defined
  in Sec. 4.2 of our related work \cite{James:2012:IE:2103656.2103667}
  has the property that its dual {{sub1}} diverges on input 0. Hence
  we could just use {{add1}} here.

\item \emph{Stop the machine.}  If we chose to stop the machine at a
  stuck state, we are essentially treating states of the form
  {{<n,0,n>}} as valid stop states for reverse execution. This also
  has the consequence of making our isomorphism total and also gives
  us two valid start states in the case of forward execution. If we
  choose this approach, the machine would look as shown below by the
  end of {{step 3}}:

@@

More generally, how the programmer handles stuck states depends on the
specific program. For some machines it may be feasible to use a
stronger type system while for others, such as our tree example in
Sec. \ref{sec:tree}, it is acceptable to diverge.

@@

The type of binary trees we use is {{@@x.(nat+x*x)}}, i.e., binary trees with
no information at the nodes and with natural numbers at the leaves. To define
the abstract machine, we need a notion of \emph{tree contexts} to track which
sub-tree is currently being explored. The definitions are shown on the left:

@@

The reduction rules on the right traverses a given tree and increments
every leaf value.  The machine here is a little richer than the ones
dealt with previously. In particular, we have a two types of machine
states -- {{<t,c>}} and {{ {[t, c]} }}. The first of these corresponds
to walking down a tree, building up the context in the process. The
second corresponds to reconstructing the tree from the context and
also switching focus to any unexplored sub-trees in the process. There
are also two syntactic categories to deal with (trees and tree
contexts) where previously we only had numbers. The {{fold}} and
{{unfold}} isomorphisms that we need for trees and tree contexts are:

@@

On the input side, for {{T}} states, we have expanded the tree component and
for {{F}} states we have expanded the tree contexts. Similarly we have done
the opposite on the output side exactly matching up what the abstract machine
does. One thing to note is that we dropped the {{1}} introduced by expanding
the {{bool}} and instead just labeled the {{true}} and {{false}} branches.

We are ready to start connecting the machine states corresponding to
the reductions that we would like:
\begin{enumerate}
\item When we encounter a leaf we would like to increment its value and move
  to the corresponding {{F}} machine state. For the sake of simplicity, in
  this interpreter we won't be concerned with exposing stuck states on
  reverse execution caused by {{sub1 0}}. So we are free to use the {{add1}}
  combinator.

\item For all the other reduction rules, it is a straightforward mapping of
  related states following the reduction rules.

@@

The required reduction rules have been labeled on the construction
belong. We have also included subscripts on the various {{t}}s to
indicate any swaps that may need to be inserted.

@@

This essentially completes the construction of the (partially)
isomorphic tree-traversal interpreter, except for the boring final
step. (Partially because of the implicit non-termination introduced by
the {{add1}}).

@@

In the final construction we present a meta-circular interpreter for
{{langRevT}} written in {{langRevT}}. This is a a non-trivial abstract
machine with several cases, but the deriving the corresponding
{{langRevT}} interpreter follows in exactly the same way as
before. Here is a logically reversible small-step abstract machine for
{{langRevT}} and the derived isomorphic interpreter with the
reductions labeled.

@@

\item The definition here shows only the handling of the composition
  combinators, which are the interesting cases. All the primitive
  isomorphisms are hidden in the `{{iso(v)}}' application in
  {{rule~1}}. This should be read as `transform {{v}} according to the
  primitive isomorphism {{iso}}'. Implementing this is straightforward
  and is hence skipped.

\item Stuck states in this machine correspond to runtime values not
  matching their expected types. They can be handled as described in
  Sec. \ref{sec:adder} using GADTs (which eliminates them entirely),
  divergence, or adding extra halting states. We have abstracted from
  this choice and marked the relevant cases with combinators labeled
  `{{cast}}'.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
% Previous work on {{langRevT}} \cite{rc2011,infeffects} studied its
% properties and various constructions in it. However, for a
% computational model, there has been little focus on actually writing
% programs directly in {{langRevT}}. This paper aims to remedy that by
% presenting a systematic technique of expressing logically reversible
% small-step abstract machines in {{langRevT}}. In other words, this
% paper shows that once we have a logically reversible machine in a
% notation of our choice, expressing the machine as an isomorphic
% interpreter in {{langRevT}} can be done systematically and does not
% present any significant conceptual difficulties. This gives us a means
% of developing large {{langRevT}} programs with the ease of reasoning
% at the level of a conventional small-step semantics.  We develop
% several simple interpreters over numbers and addition, move onto tree
% traversal and finish with an interpreter for {{langRevT}} expressed in
% {{langRevT}}.
  In our previous work, we developed a reversible programming language and
  established that every computation in it is a (partial) isomorphism that is
  reversible and that preserves information. The language is founded on type
  isomorphisms that have a clear categorical semantics but that are awkward
  as a notation for writing actual programs, especially recursive ones. This
  paper remedies this aspect by presenting a systematic technique for
  developing a large and expressive class of reversible recursive programs,
  that of logically reversible small-step abstract machines. In other words,
  this paper shows that once we have a logically reversible machine in a
  notation of our choice, expressing the machine as an isomorphic interpreter
  can be done systematically and does not present any significant conceptual
  difficulties. Concretely, we develop several simple interpreters over
  numbers and addition, move on to tree traversals, and finish with a
  meta-circular interpreter for our reversible language. Generally, this
  gives us a means of developing large reversible programs with the ease of
  reasoning at the level of a conventional small-step semantics.
\end{abstract}

@@

In this paper, we investigate the practicality of writing recursive programs
in {{langRevT}}. Specifically, we assume that we are giving some reversible
recursive algorithm expressed as a \emph{reversible abstract machine}, and we
show via a number of systematic steps, how to develop a {{langRevT}} program
from that abstract machine. Our choice of starting from reversible abstract
machines is supported by the following observations:
\begin{itemize}
\item it is an interesting enough class of reversible programs: researchers
  have invested the effort in manually designing such machines, e.g., the
  SE(M)CD machine~\cite{Kluge:1999:SEMCD}, and the SECD-H~\cite{lorenz};
\item every reversible interpreter can be realized as such a machine: this
  means that our class of programs includes self-interpreters which are
  arguably a measure of the strength of any reversible
  language~\cite{Axelsen:2011:RPC:1987171.1987176,Yokoyama:2008:PRP,Yokoyama:2007:RPL:1244381.1244404};
\item general recursive programs can be systematically transformed to
  abstract machines: the technique is independent of reversible programs and
  consists of transforming general recursion to tail recursion and then
  applying \emph{fission} to split the program into a driver and a small-step
  machine~\cite{Danvy:2008:RRN:1813347.1813350,Rendel:2010:ISD:1863523.1863525}.
\end{itemize}

@@

\paragraph*{Graphical Language.}
Following the tradition established for monoidal
categories~\cite{springerlink:10.1007/978-3-642-12821-94}, we present a
graphical language that conveys the semantics of the language.  The general
idea of the graphical notation is that combinators are modeled by ``wiring
diagrams'' or ``circuits'' and that values are modeled as ``particles'' or
``waves'' that may appear on the wires. Evaluation therefore is modeled by
the flow of waves and particles along the wires.

@@

\item Operations {{fold}} and {{unfold}} are specific to each recursive
  data type and are drawn as triangular boxes. For instance,
  Sec. \ref{sec:bounded} shows {{fold}}/{{unfold}} isomorphisms for
  numbers, {{nat=@@x.(1+x)}}.

\item The {{trace+}} operation is drawn as a looped circuit, where the
  traced type {{b1}} is shown as flowing backward. 
%% The semantics and
%%  interpretation of {{trace}} are dealt with in detail in previous
%%  work (see Sec. 4 of \cite{James:2012:IE:2103656.2103667}).

@@

Our goal is to implement this abstract machine in {{langRevT}}: We start by
writing a combinator {{c : nat * nat <-> nat * nat}} that executes one step
of the machine transition. Here {{nat}} is an abbreviation for the recursive
type {{@@x.1+x}}. The combinator, when iterated, should map {{(n,0)}} to
{{(0,n)}} by precisely mimicking the steps of the abstract machine. Let us
analyze this desired combinator {{c}} in detail starting from its interface:

@@

The triangle on the left denotes the {{unfold}} isomorphism, which works as
follows: If the number {{n}} is zero, the output is the top branch which has
type {{1}}. If the number was non-zero, the output is on the bottom branch
and has value {{n-1}}. The triangle on the right is the dual {{fold}}
isomorphism.  We now use distribution to isolate each possible machine state:

@@

At this point, it is clear that the machine's transition consists connecting
the {{n-1}} input wire to {{n'}} on the output side and the {{m}} input wire
to {{m'-1}} on the output side:

@@

We are now close to the completion of the interpreter.  The branch labeled
{{((), m)}} corresponds to the machine state {{<0,m>}} which is the stop
state of the machine. Similarly the branch labeled {{((), n')}} corresponds
to the start state of the machine.  The last step in the construction
involves introducing a {{trace}} operation for iterating the small step
realized so far. We do this by first ``sliding'' the two sections of the
diagram as shown below:

@@

This rearrangement exposes that the middle connections were really recursive
calls to the machine transition. The final figure is given below:

%% The introduced {{trace}} feeds back related machine states. The exposed start
%% state essentially requires the initial value of {{n}} to start the machine
%% and the exposed stop state returns any result values that have been computed.

@@

A closer look however reveals that the machine defines a
\emph{partial} isomorphism: not all valid final states can be mapped
to valid start states. Indeed consider the configuration {{<0,2,3>}}
which is a valid final state. Going backwards, the transitions start
as follows {{<0,2,3> |--> <1,1,2> |--> <2,0,1>}} at which point, the
machine gets stuck at a state that is not a valid start state. This 
is a general problem that we discuss below in detail.

\paragraph*{Stuck States.}
The type systems of most languages are not expressive enough to encode the
precise domain and range of a function. For example, in most typed languages,
division by zero is considered type-correct and the runtime system is
required to deal with such an error. A stuck state of an abstract machine is
just a manifestation of this general problem. The common solutions are:
\begin{itemize}
\item \emph{Use a more expressive type system.}  One could augment
  {{langRevT}} with a richer type system that distinguishes non-zero numbers
  from those that can be zero, thereby eliminating the {{sub1 0}} situation
  entirely. Similarly, in the meta-circular interpreter in
  Sec.~\ref{sec:langrevt-int}, one could use generalized abstract data types
  (GADTs)~\cite{phantom,Xi:2003:GRD:604131.604150}) to eliminate the stuck
  states.

\item \emph{Diverge.}  Another standard approach in dealing with stuck states
  is to make the machine diverge or leave the output undefined or
  unobservable in some way. If the specific case of the machine above, we can
  use a primitive {{add1}} whose dual {{sub1}} is undefined when applied to
  {{0}}. (Such a primitive is definable in
  {{langRevT}}~\cite[Sec.~4.2]{James:2012:IE:2103656.2103667}.)

\item \emph{Stop the machine.}  Alternatively, we can consider the stuck
  state as a valid final state. In the case of the example above, we would
  treat states of the form {{<n,0,n>}} as valid stop states for reverse
  execution. This gives us two valid start states in the case of forward
  execution and makes the isomorphism total. If we chose this approach, the
  machine would look as shown below by the end of {{step 3}}:

@@

% More generally, how the programmer handles stuck states depends on the
% specific program. For some machines it may be feasible to use a
% stronger type system while for others, such as our tree example in
% Sec. \ref{sec:tree}, it is acceptable to diverge.

@@

The type of binary trees we use is {{@@x.(nat+x*x)}}, i.e., binary trees with
no information at the nodes and with natural numbers at the leaves. To define
the abstract machine, we need a notion of \emph{tree contexts} to track which
subtree is currently being explored. The definitions are shown on the left:

@@

The reduction rules on the right traverse a given tree and increment
every leaf value.  The machine here is a little richer than the ones
dealt with previously. In particular, we have a two types of machine
states -- {{<t,c>}} and {{ {[t, c]} }}. The first of these corresponds
to walking down a tree, building up the context in the process. The
second corresponds to reconstructing the tree from the context and
also switching focus to any unexplored subtrees in the process. There
are also two syntactic categories to deal with (trees and tree
contexts) where previously we only had numbers. The {{fold}} and
{{unfold}} isomorphisms that we need for trees and tree contexts are:

@@

On the input side, for {{T}} states, we have expanded the tree component and
for {{F}}~states we have expanded the tree contexts. Similarly we have done
the opposite on the output side exactly matching up what the abstract machine
does. One thing to note is that we dropped the {{1}} introduced by expanding
the {{bool}} and instead just labeled the {{true}} and {{false}} branches.

We are ready to start connecting the machine states corresponding to
the reductions that we would like:
\begin{enumerate}
\item When we encounter a leaf we would like to increment its value and move
  to the corresponding {{F}} machine state. For the sake of simplicity, in
  this interpreter we won't be concerned with exposing stuck states: we
  simply use {{add1}} whose adjoint {{sub1}} diverges when applied to {{0}}.

\item For all the other reduction rules, it is a straightforward mapping of
  related states following the reduction rules. For readability, we have
  annotated the diagram below with the names of the reduction rules and we
  have included subscripts on the various {{t}}s to indicate any implicit
  swaps that should be inserted.

@@

@@

This essentially completes the construction of the (partially) isomorphic
tree-traversal interpreter, except for the final step which slides the input
and output sides.

@@

In the final construction we present a meta-circular interpreter for
{{langRevT}} written in~{{langRevT}}. This is a a non-trivial abstract
machine with several cases, but the derivation follows in exactly the same
way as before. Here is a logically reversible small-step abstract machine for
{{langRevT}} and the derived isomorphic interpreter with the reductions
labeled.

@@

\item The definition here shows only the handling of the composition
  combinators, which are the interesting cases. All the primitive
  isomorphisms are hidden in the {{iso(v)}} application in {{rule~1}}. This
  should be read as ``transform {{v}} according to the primitive isomorphism
  {{iso}}.'' Implementing this is straightforward and is hence skipped.

\item Stuck states in this machine correspond to runtime values not
  matching their expected types. They can be handled as described in
  Sec. \ref{sec:adder} using GADTs (which eliminates them entirely),
  divergence, or adding extra halting states. We have abstracted from
  this choice and marked the relevant cases with combinators labeled
  {{cast}}.
